### 2023-02-10 draft.tex
Deploying to gh-pages from @ PoisotLab/ms_interpreting_probabilities@55e45e6fe79c220764120a9d273be381050b76e1 🚀

Cataloging species interactions across space is a gargantuan task. At
the core of this challenge lies the spatiotemporal variability of
ecological networks (Poisot \emph{et al.} 2012, 2015), which makes
documenting the location and timing of interactions difficult. Indeed,
it is not sufficient to know that two species have the biological
capacity to interact to infer the realization of their interaction at a
specific time and space (Dunne 2006). Taking food webs as an example, a
predator species and its potential prey must first co-occur on the same
territory in order for a trophic interaction to take place (Blanchet
\emph{et al.} 2020). They must then encounter, which is conditional on
their relative abundances in the ecosystem and the matching of their
phenology (Poisot \emph{et al.} 2015). Finally, the interaction occurs
only if the predators have a desire to consume their prey and are able
to capture and ingest them (Pulliam 1974). Environmental (e.g.
temperature and presence of shelters) and biological (e.g.~physiological
state of both species and availability of other prey species) factors
contribute to this variability by impacting species co-occurrence
(Araujo \emph{et al.} 2011) and the realization of their interactions
(Poisot \emph{et al.} 2015). In this context, the development of
computational methods in ecology can help alleviate the colossal
sampling efforts required to document species interactions across time
and space (Strydom \emph{et al.} 2021). Having a better portrait of
species interactions and the emerging structure of their food webs is
important since it lays the groundwork for understanding the
functioning, dynamics, and resilience of ecosystems worldwide (e.g.,
Proulx \emph{et al.} 2005; Pascual \emph{et al.} 2006; Delmas \emph{et
al.} 2019).

@@

The first aspect to take into consideration when estimating or using
probabilities of interactions is knowing if they describe the likelihood
of potential or realized interactions. A potential interaction is
defined as the biological capacity of two species to interact (i.e., the
probability that they \emph{can} interact) whereas a realized
interaction refers to the materialization or observation of this
interaction in a delineated space and time period (i.e., the probability
that they interact). Here, we will use the terms \emph{metaweb} to
designate networks of potential interactions and \emph{local networks}
for those of realized interactions. Frequent confusion arises among
ecologists over the use of these two terms, especially in a
probabilistic context. Indeed, in many studies of probabilistic
ecological networks, it remains unclear when authors describe potential
or realized interactions, or when so-called probabilities are actually
\emph{interaction scores}. A better understanding of these differences
would alleviate interpretation errors and help ecologists use these
numbers more appropriately.

@@

We can express the probability that two taxa \(i\) and \(j\) can
interact as

\begin{equation}\protect\hypertarget{eq:metaweb}{}{P(i \rightarrow j),}\label{eq:metaweb}\end{equation}

@@

Starting from a selected set of species, metawebs can be built using
different data sources, including literature review, field work, and
predictive models (e.g., the metaweb of Canadian mammals inferred by
Strydom \emph{et al.} 2022). Every pair of species that has been
observed to interact at least once can be given a probability of
interaction of \(1\); we know that they \emph{can} interact. This means
that rare interactions can be given high probabilities in the metaweb.
Unobserved interactions, on the other hand, are given lower
probabilities, going as low as \(0\) for forbidden links. Two important
nuances must however be made here. Because of observation errors due to
taxonomic misidentifications and ecological misinterpretations, many
observations of interactions are actually false positives. Similarly,
forbidden interactions can be false negatives if e.g.~they have been
assessed for specific phenotypes, locations or time. Implementing a
Bayesian framework, which updates prior probabilities of interactions
with empirical data, could lessen these errors.

One potential application of metawebs is the prediction of networks of
realized interactions. Even though the spatiotemporal variation of
interactions is not considered in metawebs, they can nevertheless be a
useful tool to reconstruct local networks. First, metawebs set the upper
limit of species interactions, i.e.~the probability that two species
interact at a specific location is always lower than the probability of
their potential interaction
(\(P(i \rightarrow j) < P(i \rightarrow j | x_1, x_2, ...)\)). This
implies that predicting local networks from metawebs realistically might
need be done by filtering out interactions from its subnetworks. Second,
the structure of local networks is constrained by the one of their
metaweb. Inferring their structure from the metaweb first could help
predicting which potential interactions are not realized. Ultimately, we
believe that predicting probabilistic local networks from metawebs must
be done with caution, as long as uncertainty is adequately propagated.

@@

\[P(i \rightarrow j | C, A, N, E, t)\]

i = predator j = prey C = co-occurrence (and relative abundance?) A =
area N = relative abundance (or network?) E = environment (including
network?) t = time

What does a probability in the context of a local network mean? A
cautionary tale of how we define probabilities.

\begin{itemize}
\tightlist
\item
  It means the probability that two taxa will interact at a given
  location.
\item
  What do we mean by saying that two taxa will interact? We usually mean
  that at least one individual from one group will interact with (e.g.,
  predate) at least one other individual from the other group.
\item
  The probability is conditional on the environmental and local
  abundance contexts.
\item
  We should expect a certain number of interactions to be realized
  depending on the probability value. This number depends on the number
  of trials, which also depends on the ecological context (e.g.,
  environmental conditions, scale) in which probabilities were
  estimated. This is in contrast with probabilities in metawebs.
\end{itemize}

@@

@@

@@

>>>>

Cataloging species interactions across space is a gargantuan task. At
the core of this challenge lies the spatiotemporal variability of
ecological networks (Poisot \emph{et al.} 2012, 2015), which makes
documenting the location and timing of interactions difficult. Indeed,
it is not sufficient to know that two species have the biological
capacity to interact to infer the realization of their interaction at a
specific time and space (Dunne 2006). Taking food webs as an example, a
predator species and its potential prey must first co-occur in order for
a trophic interaction to take place (Blanchet \emph{et al.} 2020). They
must then encounter, which is conditional on their relative abundances
in the ecosystem and the matching of their phenology (Poisot \emph{et
al.} 2015). Finally, the interaction occurs only if the predators have a
desire to consume their prey and are able to capture and ingest them
(Pulliam 1974). Environmental (e.g.~temperature and presence of
shelters) and biological (e.g.~physiological state of both species and
availability of other prey species) factors contribute to this
variability by impacting species co-occurrence (Araujo \emph{et al.}
2011) and the realization of their interactions (Poisot \emph{et al.}
2015). In this context, the development of computational methods in
ecology can help alleviate the colossal sampling efforts required to
document species interactions across time and space (Strydom \emph{et
al.} 2021). Having a better portrait of species interactions and the
emerging structure of their food webs is important since it lays the
groundwork for understanding the functioning, dynamics, and resilience
of ecosystems worldwide (e.g., Proulx \emph{et al.} 2005; Pascual
\emph{et al.} 2006; Delmas \emph{et al.} 2019).

@@

The first aspect to take into consideration when estimating or
interpreting probabilities of interactions is knowing if they describe
the likelihood of potential or realized interactions. A potential
interaction is defined as the biological capacity of two species to
interact (i.e., the probability that they \emph{can} interact) whereas a
realized interaction refers to the materialization or observation of
this interaction in a delineated space and time period (i.e., the
probability that they interact). Here, we will use the terms
\emph{metaweb} to designate networks of potential interactions and
\emph{local networks} for those of realized interactions. Frequent
confusion arises among ecologists over the use of these two terms,
especially in a probabilistic context. Indeed, in many studies of
probabilistic ecological networks, it remains unclear when authors
describe potential or realized interactions, or when so-called
probabilities are actually \emph{interaction scores}. Likewise,
probabilistic potential interactions are often used as realized
interactions (and conversely), even when the type of interaction is
clearly indicated. We believe that a better understanding of these
differences and concepts would alleviate interpretation errors and help
ecologists use these numbers more appropriately.

@@

We can express the probability that two taxa \(i\) and \(j\) can
interact in a metaweb \(M\) as

\begin{equation}\protect\hypertarget{eq:metaweb}{}{P_{M}(i \rightarrow j),}\label{eq:metaweb}\end{equation}

@@

Starting from a selected set of species, metawebs can be built using
different data sources, including literature review, field work, and
predictive models (e.g., the metaweb of Canadian mammals inferred by
Strydom \emph{et al.} 2022). Every pair of species that has been
observed to interact at least once can be given a probability of
interaction of \(1\); we know that they \emph{can} interact. This means
that rare interactions can technically be given high probabilities in
the metaweb. Unobserved interactions, on the other hand, are given lower
probabilities, going as low as \(0\) for forbidden links. Two important
nuances must however be made here. Because of observation errors due to
taxonomic misidentifications and ecological misinterpretations, many
observations of interactions are actually false positives. Similarly,
forbidden interactions can be false negatives if e.g.~they have been
assessed for specific phenotypes, locations or time. Implementing a
Bayesian framework, which updates prior probabilities of interactions
with empirical data, could lessen these errors.

@@

As opposed to metawebs, probabilistic local food webs represent the
likelihood that two species will interact at a specific location and
within a given time period; in other words, they are context dependant.
To be specific, space is defined here as the geographic coordinates (\$
\bold x,
\bold y\() of the spatial boundaries delineating the system, whereas time is the time interval (\)
\bold t\$) during which interactions were sampled or for which they were
predicted. We put these variables in bold to emphasize that they are not
single values, but rather continued dimensions that could be outlined
differently depending on the study system. Regardless of how they were
defined, they always delineate a specific area \(A\) and duration \(D\).
These could refer to the sampled area and duration or to the targeted
location and time period.

Many factors could be taken into consideration when estimating the
probability that a predator species \(i\) interacts with a given prey
species \(j\) locally. One of the most important is species
co-occurrence, which describes if both species can be found at location
and time (\$ \bold x, \bold y, \bold t\$). Surely, the probability that
the interaction is realized must be \(0\) when species do not co-occur.
Similarly, species relative abundance is another important predictor of
the probability of interaction, because it impacts the probability that
species will randomly encounter (Canard \emph{et al.} 2012; Canard
\emph{et al.} 2014; Poisot \emph{et al.} 2015). Interaction
probabilities can also be conditional on other biological and
environmental variables, such as temperature, precipitation, presence of
shelters, phenotypic plasticity, phenology, and presence of other
interacting species in the network. These conditions can affect species
traits, which greatly impact the likelihood of interactions (Poisot
\emph{et al.} 2015). When probabilities of interactions are estimated
using specific values of these variables, ecologists should make them
explicit in their metadata, preferably using mathematical equations to
avoid any ambiguity. In the next section, we will see examples of why
this matters when it comes to spatial, temporal, and taxonomic scaling
of biotic interactions. Here, we will use the variable \(\bold \Omega\)
as a substitute for the biological and ecological context in which
interaction probabilities were estimated. This gives us the following
equation for the probability that species (or taxa) \(i\) and \(j\)
interact in a local network \(N\):

\begin{equation}\protect\hypertarget{eq:local}{}{P_N(i \rightarrow j | \bold x, \bold y, \bold t, \bold \Omega).}\label{eq:local}\end{equation}

\hypertarget{from-potential-to-realized-interactions}{%
\subsection{From potential to realized
interactions}\label{from-potential-to-realized-interactions}}

Even though the spatiotemporal variation of interactions is not
considered in metawebs, they can still be useful to reconstruct local
networks of realized interactions. Indeed, local networks are formed
from subsets of their metaweb (called subnetworks), which are obtained
by selecting a subset of both species and interactions (Dunne 2006).
Because a community's composition is arguably easier to sample (or
predict) than its interactions, the biggest challenge is to sample links
from the metaweb. This becomes a conceptual issue when we consider how
potential and realized interactions differ. Despite these concerns,
metawebs remain an important source of ecological information that can
be leveraged for realistically predicting spatially explicit networks.
First, metawebs set the upper limit of species interactions (McLeod
\emph{et al.} 2021), i.e. the probability that two species interact at a
specific location is always lower or equal to the probability of their
potential interaction:

\begin{equation}\protect\hypertarget{eq:switch}{}{P_N(i \rightarrow j | \bold x, \bold y, \bold t, \bold \Omega) \le P_M(i
\rightarrow j).}\label{eq:switch}\end{equation}

Therefore, building local networks from their metaweb keeping the same
values of interaction probability would generate systematic biases in
the prediction. Second, the structure of local networks is constrained
by the one of their metaweb (Saravia \emph{et al.} 2022). This suggests
that a metaweb not only constraints the pairwise interactions of its
corresponding local networks, but also their emerging properties.
Inferring the structure of local networks from the metaweb could thus
help estimate more realistically the likelihood that potential
interactions are realized and observed locally (Strydom \emph{et al.}
2021).

@@

\begin{itemize}
\tightlist
\item
  We should expect a certain number of interactions to be realized
  depending on the probability value. This number depends on the number
  of trials, which also depends on the ecological context (e.g.,
  environmental conditions, scale) in which probabilities were
  estimated. This is in contrast with probabilities in metawebs.
\end{itemize}

@@

\leavevmode\hypertarget{ref-Canard2014EmpEva}{}%
Canard, E.F., Mouquet, N., Mouillot, D., Stanko, M., Miklisova, D. \&
Gravel, D. (2014). Empirical Evaluation of Neutral Interactions in
Host-Parasite Networks. \emph{The American Naturalist}, 183, 468--479.

\leavevmode\hypertarget{ref-Canard2012EmeStr}{}%
Canard, E., Mouquet, N., Marescot, L., Gaston, K.J., Gravel, D. \&
Mouillot, D. (2012). Emergence of Structural Patterns in Neutral Trophic
Networks. \emph{PLOS ONE}, 7, e38295.

@@

\leavevmode\hypertarget{ref-Saravia2022EcoNet}{}%
Saravia, L.A., Marina, T.I., Kristensen, N.P., De Troch, M. \& Momo,
F.R. (2022). Ecological network assembly: How the regional metaweb
influences local food webs. \emph{Journal of Animal Ecology}, 91,
630--642.
